---
title: 'A data analyst workflow, part 1: SQL & tidyverse'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

As a data professional, I have enjoyed learning and using multiple tools for my workflows. For me, everything used to begin and end with R. Today, SQL is a must-know. Not being able to pull your own custom tables from a warehouse can make things tricky. Then there is `tidyverse`, the master collection of packages for data science & analytics. As an OG R user, I cannot envision data work without `tidyverse`.

In this article, I want to explore how a data analyst can use SQL and `tidyverse` for their workflow. In this first part of a 2-part article, I want to demonstrate how a data analyst can use _one OR the other for the initial stages of data exploration_, and then double down on `tidyverse`, leveraging `ggplot2` for a deeper exploration. By no means does this preclude the extensive use of SQL for data wrangling. Rather, this post showcases the wonders of `tidyverse` (a [collection](https://www.tidyverse.org/) of R packages designed for data science, sharing an underlying design philosophy, grammar, and data structures) and specifically, `ggplot2` (the [language](https://ggplot2-book.org/) of elegant graphics) for a SQL user's benefit.

## 1. The dataset and the goal

The dataset I am using is clinical. Sourced from the UCI machine learning repo, it is the *Diabetes 130-US hospitals for years 1999-2008 Data Set*. The dataset is large, ~100K rows and 51 columns in its raw format. It is, however, clean data. For the purpose of this article, in order to show SQL and `tidyverse` language in tandem, I will split it up into 5 parts, and we will assume that the data is actually available to us in these 5 different pieces, rather than as the whole, cleaned data, since this is typically the case in real life. 

I will skip the portion about [`dbplyr`](https://dbplyr.tidyverse.org/articles/dbplyr.html), referring readers to the hyperlinked article that will show you how to actually pull data from a remote database using `tidyverse`'s `dbplyr`. Typically, this is done using SQL, but`dbplyr` allows you to do this within `R`. Rather, I will focus on _the initial stages of data exploration_, using both SQL and `tidyverse` for the same output, while extending the `tidyverse` portion to include `ggplot2` visualization examples, using different plot types for each use case. Note that in each case, you can use SQL first, and then use the SQL output as an input for the `ggplot2` visualization. 

## 2. Reading in the data

The data has been split into 5 parts -- demographic, medical, hospital visits, outcome, test results. To learn more about the actual data, see [here](https://www.hindawi.com/journals/bmri/2014/781670/). Each part is connected with the other through a UID that is a concatenation of the patient encounter ID and the patient number (using either one doesn't work to make the ID unique). Note that all of the analyses in this post will be done at the UID level, rather than patient level. 

```{r diabetes, message=FALSE, warning=FALSE}
rm(list=ls())
library(sqldf)
library(dplyr)
library(readxl)
#library(dbplyr)
library(ggplot2)

dem <- read_excel("/Users/vvachharajani/Downloads/dataset_diabetes/diabetic_data.xlsx", "demo")
meds <- read_excel("/Users/vvachharajani/Downloads/dataset_diabetes/diabetic_data.xlsx", "medications")
visits <- read_excel("/Users/vvachharajani/Downloads/dataset_diabetes/diabetic_data.xlsx", "hosp_visits")
y <- read_excel("/Users/vvachharajani/Downloads/dataset_diabetes/diabetic_data.xlsx", "readmissions")
results <- read_excel("/Users/vvachharajani/Downloads/dataset_diabetes/diabetic_data.xlsx", "test_results")
```

## 3. Early explorations

Let's begin using SQL and `tidyverse` to answer some initial questions related to the dataset. The primary hypothesis for this data is the **impact of HbA1c measurement on readmission rates**, where "readmission" is our response. We will also answer a number of other questions along the way to understand the data better, using `ggplot2` when we can.

### 3.1 Look at the data

#### 3.1.1 Get some counts

Let's take a look at medications and get a sample size for it, first using SQL and then `R`. 

```{r meds}
sqldf('SELECT * FROM meds where 1=0') # SQL see col names
sqldf('SELECT uid, metformin, repaglinide, nateglinide, chlorpropamide FROM meds LIMIT 5') # SQL
head(meds, n=5) # dplyr
```

```{r sample}
sqldf('SELECT COUNT(uid) FROM meds') # SQL
nrow(meds) # R
```

How many patients with a diabetes diagnosis, vs respiratory, circulatory, etc.?

```{r diab}
sqldf('SELECT primary_diag, COUNT(*) FROM results GROUP BY primary_diag') # SQL
results %>% group_by(primary_diag) %>% count(primary_diag) # R
```

How many women came in through an emergency admission type?
```{r gender}
sqldf('SELECT gender, admission_type_id, COUNT(*) AS n FROM dem LEFT JOIN visits USING(uid) WHERE admission_type_id=1 GROUP BY gender')  # SQL
visits %>% left_join(dem, by=join_by(uid)) %>% subset(admission_type_id==1) %>% count(gender, admission_type_id) # dplyr
```

#### 3.1.2 A mosaic plot 

Instead of extracting counts manually, let's use a mosaic plot to get a sense of how 2 count variables are distributed relative to each other. In this case, age and admission type. This plot sheds light into data availability and assymetric distributions. For example, here, we see that most patients come from emergency, urgent care, or as an elective, and that there is missing or "not available" admission type data. It is important to retain these 2 categories separately, since they mean different things. Note that in the `ggplot` parameters, I have not yet introduced axes label cleanup, etc.

```{r mosaic, message=FALSE, warning=FALSE}
p0 <- dem %>% left_join(visits, by=join_by(uid)) %>%  
  mutate(admission_type=ifelse(admission_type_id==1, "1:Emergency", 
                               ifelse(admission_type_id==2, "2:Urgent", 
                               ifelse(admission_type_id==3, "3:Elective", 
                               ifelse(admission_type_id==4, "4:Newborn", 
                               ifelse(admission_type_id==5, "5:Not Available",
                               ifelse(admission_type_id==6, "6:NULL", 
                               ifelse(admission_type_id==7, "7:Trauma Center", 
                                      "8:Not Mapped")))))))) %>% 
  group_by(admission_type, age) %>% summarise(n=n()) %>% mutate(freq = n / sum(n)) 
ggplot(p0, aes(x=age, y=admission_type)) +
  geom_tile(aes(fill=n)) + scale_fill_gradient(low="white", high="blue")
```


#### 3.1.3 A simple join

Let's join all 5 datasets and look at it. Note that in SQL, in order to look only at the first few columns, we need to know the column names, which is what we first do here.


```{r join-1, results='hide'}
# (Output suppressed)
sqldf('SELECT * FROM dem LEFT JOIN visits USING(uid) LEFT JOIN results USING(uid) LEFT JOIN meds USING(uid) LEFT JOIN y USING(uid) where 1=0') # SQL
```

```{r join-2}
sqldf('SELECT uid, race, gender, age, weight FROM dem LEFT JOIN visits USING(uid) LEFT JOIN results USING(uid) LEFT JOIN meds USING(uid) LEFT JOIN y USING(uid) LIMIT 5') # SQL
dem %>% left_join(visits, by=join_by(uid)) %>% left_join(results, by=join_by(uid)) %>% left_join(meds, by=join_by(uid)) %>% left_join(y, by=join_by(uid)) %>% print(n=5) # dplyr, by default shows 10 rows, so we ask it to print 5
```

### 3.2 Explore the response: readmissions

#### 3.2.1 Lab procedures 

Let's start with the simplest question -- for the primary response variable, "readmitted", how many lab procedures were done by each category of the response? Note here that "number of lab procedures" is one of a handful of continuous design covariate -- rest of the ~45 covariates are all categorical/discrete.

```{r explore-1-1, message=FALSE, warning=FALSE}
# How many lab tests performed for readmitted patients
sqldf('SELECT readmitted, SUM(num_lab_procedures) AS n FROM visits LEFT JOIN y USING(uid) GROUP BY readmitted') # SQL
visits %>% left_join(y, by=join_by(uid)) %>% group_by(readmitted) %>% 
  summarise(n=sum(num_lab_procedures)) %>% mutate(freq = n / sum(n)) # dplyr, w/ an added proportion 
```

Since the above doesn't really tell us much, other than actual counts, proportions by response categories, let's use `ggplot2` to explore the distribution of "number of lab procedures", using a barplot/histogram approach, with "readmitted" as the `fill` element. This helps us get a better picture of their relationship; we see here how, for a strikingly normally distributed "number of lab procedures" (other than 1 outlier), on average, the higher the volume of procedures, the more the proportion of readmitted. 

```{r explore-1-2, message=FALSE, warning=FALSE}
# How many lab tests performed for readmitted patients, use ggplot2
p1 <- visits %>% left_join(y, by=join_by(uid))
ggplot(data = p1 ,aes(x=num_lab_procedures,fill=readmitted)) + geom_bar() + labs(x="Number of lab procedures", y="counts") + scale_y_continuous(
    labels = function(n) scales::comma(abs(n)))
```

Let's also do this using `ggplot`'s beautiful density plots. It is a slightly different type of visual, and tells us how the distribution of X shifts left or right by the response or `fill`.

```{r explore-1-3, message=FALSE, warning=FALSE}
ggplot(p1, aes(num_lab_procedures)) + geom_density(aes(fill=factor(readmitted)), alpha=0.8) + labs(x="Number of lab procedures") 
```

#### 3.2.2 Demographics

Next, we ask how readmissions differ across age groups and gender. Let's also plot this to understand the output better. We first use a population pyramid approach to get the counts and then barplot the proportions to get a better understanding of the variance in readmissions across these groups. 

```{r explore-2, message=FALSE, warning=FALSE}
# What are readmission rates by the different age groups?
sqldf('SELECT age, readmitted, COUNT(*) AS n FROM dem LEFT JOIN y USING(uid) GROUP BY age')  # SQL
p21 <- dem %>% left_join(y, by=join_by(uid)) %>% group_by(gender, age, readmitted) %>% summarise(n=n()) %>%
mutate(pct = 100 * n / sum(n), readmission=ifelse(readmitted=="NO", "not readmitted", "readmitted")) %>% 
  ungroup() %>% subset(gender=="Male"|gender=="Female") %>%
  ggplot() +
  geom_col(aes(x = ifelse(readmission == "readmitted", -n, n),
               y = age,
               fill = readmission)) +
  facet_wrap(~ gender) +
  scale_x_continuous(
    labels = function(n) scales::comma(abs(n))) +
  xlab("Counts")
p21
p22 <- dem %>% left_join(y, by=join_by(uid)) %>% group_by(gender, age, readmitted) %>% summarise(n=n()) %>% mutate(freq = n / sum(n)) %>% subset(gender=="Male"|gender=="Female")
ggplot(data=p22, aes(x=age, y=freq, fill=readmitted)) + geom_col() + facet_wrap(~ gender) + labs(y="proportions") + geom_text(aes(label = paste0(round(freq, 4) * 100, "%")), position = position_stack(vjust = 0.5), size=2.5, angle=90) + theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1))
```

The population pyramid is an intriguing plot type, and already tells us that for most age groups, more women are readmitted. But this could be solely because there are more women than men in the sample. However, from the proportion barchart, we see here that proportion of readmitted women is greater than men, particularly for the 20-30 age group.

#### 3.2.3 Patient diagnoses 

Finally, how are readmission rates distributed by patient and patient care features. For example, how is it distributed by patient primary diagnosis? In the final section of this post, we will leverage `ggplot2`'s visualization power to triangulate patient diagnoses with the key covariate and the response. Like in the previous section, we use proportions, adding the relevant labels to more easily infer that we see higher readmission rates for a diabetes diagnosis. 

We change around quite a few of the plotting parameters in `ggplot2` to make it look much more eye-catching. 

```{r explore-3, message=FALSE, warning=FALSE}
# How is readmitted rate distributed by diagnoses?
sqldf('SELECT primary_diag, readmitted, COUNT(*) as n FROM results LEFT JOIN y USING(uid) GROUP BY primary_diag, readmitted')  # SQL
p3 <- results %>% left_join(y, by=join_by(uid)) %>% group_by(primary_diag, readmitted) %>% summarise(n=n()) %>% mutate(freq = n / sum(n))
ggplot(data=p3, aes(x=primary_diag, y=n, fill=readmitted)) + geom_bar(position = "dodge", stat = "identity", color="black") + scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9")) + theme_minimal() + labs(x="Primary diagnoses", y="Counts (proportions as labels)") + geom_text(aes(label = paste0(round(freq, 4) * 100, "%")), position = position_dodge(width = 1), vjust=-0.7, size=3) + scale_y_continuous(labels = function(n) scales::comma(abs(n)))
```

#### 3.2.4 HbA1c measurement

One of the key questions this dataset seeks to answer is the *impact of the A1C test (decision to test) on readmission rates*, in the presence of covariates (especially the primary diagnosis). Output in its raw form (i.e. untransformed) doesn't always give us the answer clearly. To get around this, we will use `CASE WHEN` in SQL and `mutate` in `tidyverse`. 

Let's plot this in 2 ways -- a barplot with labels, and a spineplot. The latter allows us to see the "weight" of the underlying categories. 

```{r explore-4, message=FALSE, warning=FALSE}
# What is the readmission rate profile of patients who had their A1C measured?
sqldf('SELECT CASE WHEN A1Cresult = "None" THEN "not measured" ELSE "measured" END AS a1c, readmitted,
   COUNT(*) FROM results LEFT JOIN y USING(uid) GROUP BY a1c, readmitted ORDER BY 1') # SQL
p4 <- results %>% left_join(y, by=join_by(uid)) %>% mutate(a1c=ifelse(A1Cresult=="None", "not measured", "measured")) %>% 
  group_by(a1c, readmitted) %>% summarise(n=n()) %>% mutate(freq = n / sum(n)) 
ggplot(data=p4, aes(x=a1c, y=freq, fill=readmitted)) + geom_col() + labs(x="HbA1c test measurement", y="proportions") + geom_text(aes(label = paste0(round(freq, 4) * 100, "%")), position = position_stack(vjust = 0.5), size=3)# dplyr
# Spineplot
library(ggmosaic)
p5 <- results %>% left_join(y, by=join_by(uid)) %>% left_join(dem, by=join_by(uid))%>% mutate(a1c=ifelse(A1Cresult=="None", "not measured", "measured")) %>% subset(gender=="Male"|gender=="Female")
per <- p5 %>% group_by(a1c, readmitted) %>% summarise(n=n()) %>% mutate(freq = n / sum(n)) 
g <- ggplot(p5) + geom_mosaic(aes(x = product(a1c),fill = readmitted)) 

g + geom_text(data = ggplot_build(g)$data[[1]] %>% 
                group_by(x__a1c) %>%
                mutate(pct = .wt/sum(.wt)*100), 
              aes(x = (xmin+xmax)/2, y = (ymin+ymax)/2, label=paste0(round(pct, 2), "%")))

```

We observe a lower readmission rate (<30 days) when there is an A1C measurement taken, vs when it is not measured at all. In the 2nd/spineplot, we see this without actually calculating the percentages, while also inferring that number of patients not measured is much higher than those measured. We do however, manually add in the percentages to the spineplot to get a more complete picture on the relationship between HbA1c measurement and readmission rates. 
These are key findings which we will explore in greater detail, using `tidyverse` and `ggplot2` more extensively, in the next part of this blog series, including cutting these plots across multiple covariates to explore how HbA1c affects readmissions in the presence of other patient groupings. Stay tuned!  


